---
title: "Project 1: Covid-19 Cases and Testing"
author: "Alphoncina (Allie) Lyamuya"
date: "8/27/2021"
output:
  html_document:
    toc: yes
    theme: lumen
    toc_float: yes
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction 

Since it was first reported in Wuhan, China in December 2019, Covid-19 has
dramatically changed the world as we know it. From shutdowns across the globe,
most of which began earlier in March 2020, Covid-19 has not only devastated the
global economy but also claimed so many lives across continents. In efforts to
curb down the devasting effects of this "unprecedented global crisis",
countries, international organizations, and institutions across the world have
adopted varied measures to mitigate the effects of Covid-19. It is important to
recognize that different countries were hit differently by Covid-19. While some
had many cases of Covid-19 infections and record high in Covid-related deaths,
some had relatively less positive cases and mortality rates. Even efforts to
"flatten the curve" were not the same across countries. In this analysis, I'm
interested to answer the question: Which countries have had the highest number
of positive cases against the number of tests? I will be using the [dataset from
Kaggle](https://www.kaggle.com/lin0li/covid19testing) which include data
collected between the 20th of January and the 1st of June 2020.

***


## Data Importation and Exploration 

```{r warning=FALSE}
# Load libraries 
library(readr)
library(tidyverse)
library(psych)
library(ggplot2)
```


```{r}
# Import the data
covid_df <- read_csv("covid19.csv")
covid_df1 <- covid_df

# Explore the dataset
dim(covid_df)
vector_cols <- colnames(covid_df)
vector_cols

head(covid_df)
glimpse(covid_df)

```


***

## Data Cleaning and Transformation

```{r}
# Filtering relevant observations
covid_df_all_states <- covid_df %>%
  filter(Province_State == "All States") %>%
  select(-Province_State)
```


```{r}
# Selecting relevant variables 
covid_df_all_states_daily <- covid_df_all_states %>%
  select(Date, Country_Region, active, hospitalizedCurr, daily_tested, 
         daily_positive)

```

```{r}
# Transform the dataset
covid_df_all_states_daily_sum <- covid_df_all_states_daily %>%
  group_by(Country_Region) %>%
  summarise(
    tested = sum(daily_tested),
    positive = sum(daily_positive),
    active = sum(active),
    hospitalized = sum(hospitalizedCurr)
  ) %>%
  arrange(-tested)

covid_df_all_states_daily_sum

# Extract the top 10 rowa from the dataset
covid_top_10 <- head(covid_df_all_states_daily_sum, 10)
covid_top_10

```


***

## Analyze the Dataset

```{r}
# Creating vectors
countries <- covid_top_10$Country_Region
tested_cases <- covid_top_10$tested
positive_cases <- covid_top_10$positive
active_cases <- covid_top_10$active
hospitalized_cases <- covid_top_10$hospitalized

# Name the vectors with countries' name vector
names(tested_cases) <- countries
names(positive_cases) <- countries 
names(active_cases) <- countries 
names(hospitalized_cases) <- countries

# Identify top 3 positive against tested cases 
positive_cases / tested_cases

positive_tested_top_3 <- c(0.113260617, 0.108618191, 0.080711720)
names(positive_tested_top_3) <- c("United Kingdom", "United States", "Turkey")

positive_tested_top_3
```

***

```{r}
# Create vectors
united_kingdom <- c(0.11, 1473672, 166909, 0, 0)
united_states <- c(0.10, 17282363, 1877179, 0, 0)
turkey <- c(0.08, 2031192, 163941, 2980960, 0)

# Create matrix 
covid_mat <- rbind(united_kingdom, united_states, turkey)

# Rename the cols of the matrix
colnames(covid_mat) <- c("Ratio", "tested", "positive", "active", 
                         "hospitalized")

covid_mat
```

***

```{r}
# Create a character variable 
question <- "Which countries have had the highest number of positive cases 
against the number of tests?"

answer <- c("Positive tested cases" = positive_tested_top_3)

# Create list
dataframes_list <- list(covid_df, covid_df_all_states, 
                        covid_df_all_states_daily, covid_top_10)
matrices_list <- list(covid_mat)
vectors_list <- list(vector_cols, countries)

data_structure_list <- list(dataframes_list, matrices_list, vectors_list)

# Create another list
covid_analysis_list <- list(question, answer, data_structure_list)
covid_analysis_list
```


***

## Discussion and Conclusion 

In this analysis, I was interested to answer the question:  Which countries have
had the highest number of positive cases against the number of tests? To answer
this question, I first filtered the dataset by "All States" to explore the
values based on Covid-19 data by individual countries. I then selected
variables--columns related to daily measures--needed to answer the
aforementioned question. The dataset was then grouped by country names and
summarized by summing the values of each measure in the dataset. Since the
question is seeking the highest number of positive cases against the number of
tests, the top 10 countries were extracted from the dataset for further
analysis. 

From the top 10 data, the top 3 countries were identified by
calculating the ratio of positive cases over the number of tests. The results of
the top three countries and their ratios were stored in a vector named
"positive_tested_top_3". To make the results of this analysis more accessible
and digestible, several lists and lists of lists were created. All these lists
were stored in the list "covid_analysis_list" which can be used to explore all
the results leading up to the answer to the question explored in the analysis.








