---
title: "DACSS 601 Final Project"
author: "Alphoncina Lyamuya"
date: "8/7/2021"
output:
  pdf_document:
    toc: yes
  html_document:
    toc: yes
    theme: lumen
    toc_float: yes
subtitle: Global Disparities in Internet Access
bibliography: ref.bib
biblio-style: apalike
link-citations: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Internet access is one of the 21st-century essentials to almost all spheres of
life. Despite the push to make the internet a public utility in most advanced
economies as well as the 2016 UN General Assembly passage of a non-binding
Resolution declaring "Internet access a human right", Internet access has
remained uneven across the globe [@un_general_assembly_promotion_2016;
@brendan_net-neutrality_2014]. The International Telecommunication Union (ITU)
reported that by the end of 2016, there were 3.9 billion people that were not
using the Internet [@international_telecommunication_union_ict_2016]. With a
total population of 7.4 billion in 2016, this implies about 53% of the world's
population was digitally excluded [@world_bank_world_nodate;
@international_telecommunication_union_ict_2016]. When a significant portion of
the population of the Global North countries has been reported as active
Internet users, many in the Global South are digitally excluded. For instance,
when only 21% of the European population was reported as non-users in 2016,
about 75% of the population in Africa was still not using the Internet
[@international_telecommunication_union_ict_2016].

Most people access the internet through a certain kind of broadband connection.
Also referred to as high-speed internet access, broadband includes the following
types of connections: DSL (Digital Subscriber Line), fiber-optic, cable,
satellite, and wireless [@federal_communications_commission_types_2014;
@international_telecommunication_union_birth_2003)]. When DSL, fiber-optic,
cable and satellite connections are categorized as fixed broadband, "wireless
broadband can be mobile or fixed"
[@federal_communications_commission_types_2014]. By 2017, there were about 7.75
billion mobile broadband subscriptions and about 1.02 billion fixed broadband
subscriptions [@world_bank_world_nodate]. With the Covid-19 pandemic and the
ensued global shutdowns, access to reliable high-speed Internet has become more
than ever important. From education, to health services, to jobs and the economy
at large, fixed broadband internet has emerged as the cornerstone to social
interactions, economic functionality, and even public health responses. But this
has largely been relevant to those that can afford the cost of high-speed
internet and live in regions or places with functioning Internet infrastructure.
Given the growing need for high-speed Internet, the existing disparities in
Internet access, and the socio-economic potential of fixed broadband connection,
this study asks: to what extent does a country's per capita GDP corresponds to
access to fixed broadband internet? I hypothesized that countries with higher
per capita GDP are more likely to have well-established internet physical
infrastructure as well as high access to percapelectity and high median income
which allow most of their population to have access to fixed broadband Internet.
The findings from this study would be useful in informing international
policymakers and stakeholders in their effort to design effective measures to
address different layers of the global digital divide.


***

## Data 

The dataset used for this project is a compilation of several secondary datasets
all from the [World Bank Open Data](https://data.worldbank.org). These datasets
include data on fixed broadband subscriptions from 177 countries for the year
2007 to 2017; data on per capita GDP from 177 countries for the year 2007 to
2017; data on percapelectity access from 177 countries for the year 2007 to
2017; data on population from 177 countries for the year 2007 to 2017; and data
on cellular mobile subscriptions from 177 countries for the year 2007 to 2017.
The compiled dataset included the following variable: *country* (name of the
country), *year* (period covered in the data set, 2007-2017), *population*
(total population of a country in a given year), *fixbroads* (fixed broadband
subscriptions in a given year), *percapelect* (percapelectity access across a
country in a given year), *gdpercap* (per capita GDP of a country in a given
year), *mobsub* (cellular mobile subscriptions in a country). This dataset was
then analyzed using R statistical software to examine the statistical
relationship between the dependent variable (fixed broadband subscriptions) and
independent variable (per capita GDP).


### Import Data and Run Libraries

```{r echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
library(readr)
library(tidyverse)
library(tidyr)
library(ggplot2)
library(psych)
```

<br>

```{r}
# Import the dataset 
myData <- read.csv("newData.csv")
myData1 <- myData

# Snapshot of the dataset
head(myData)

# Description of the data
glimpse(myData)

```

### Variable Description 

The dependent variable (DV) fixed broadband subscription (`fixbroads`) represent
total residential and organizational subscriptions to high-speed access to the
Internet "at downstream speeds equal to, or greater than, 256 kbit/s. This
includes cable modem, DSL, fiber-to-the-home/building, other fixed
(wired)-broadband subscriptions, satellite broadband" (`World bank`).

The independent variable (IV) per capita GDP (`gdpercap`) represents the gross
domestic product per midyear population of a country in a given year.


```{r}
# Describing the variable 
str(myData$fixbroads)
str(myData$gdpercap)

summary(myData$fixbroads)
summary(myData$gdpercap)

```


### Data Cleaning and Transformation 

```{r}
# Check for missing values 
table(is.na(myData))

```

<br>

```{r echo=TRUE, results='hide'}
# Examine the variables with missing values
myData[rowSums(is.na(myData)) > 0, ]

table(is.na(myData$aliteracy))
```

<br>

* There are **103** missing values in the dataset `myData`. These missing values
are found across `fixbroads`, `gdpercap`, `mobsub`, and `percapelect`. To avoid
drawing inaccurate inference about the data, and given their relatively smaller
size compared to the total values in the dataset, I will not remove existing
missing values from the dataset.


### Data Transformation

Here some variable will be renamed, some new ones will be created, and the
dataset will be arranged by names of countries (`country`).

```{r}
# Transforming variables
myDataNew <- myData %>%
  rename("country" = "country_name") %>%
  mutate(percapfbroad = fixbroads/population, 
         percapmobsub = mobsub/population,
         percapelect = electric/population) %>%
  arrange(country)

head(myDataNew)

```


```{r}
# Filtering data by year
for (year in myDataNew){
  filter2007 <- filter(myDataNew, year == 2007)
  filter2008 <- filter(myDataNew, year == 2008)
  filter2009 <- filter(myDataNew, year == 2009)
  filter2010 <- filter(myDataNew, year == 2010)
  filter2011 <- filter(myDataNew, year == 2011)
  filter2012 <- filter(myDataNew, year == 2012)
  filter2013 <- filter(myDataNew, year == 2013)
  filter2014 <- filter(myDataNew, year == 2014)
  filter2015 <- filter(myDataNew, year == 2015)
  filter2016 <- filter(myDataNew, year == 2016)
  filter2017 <- filter(myDataNew, year == 2017)
} 

```


### Summary Description and Data Exploration

After creating a variable `percapfbroad` (fixed broadband subscription (per 100
people)) from the variable `fixbroads`, `percapfbroad` will be used as a DV
throughout the rest of the analysis.

```{r}
options(scipen = 999, digits = 3)

# Exploration of measures of central tendencies and distribution of the variable
describe(myDataNew$percapfbroad)
describe(myDataNew$gdpercap)

# Visual distribution of the variables
plot1a <- ggplot(data = myDataNew, mapping = aes(x = percapfbroad)) +
  geom_histogram(fill = "chartreuse2", bins = 15, color = "white") +
  labs(title = "Per Capita Fixed Broadband 
       Subscriptions for 2007-2017",
     caption = "Source:https://data.worldbank.org") 

plot1b <- ggplot(data = myDataNew, mapping = aes(x = gdpercap)) +
  geom_histogram(fill = "bisque3", bins = 15, color = "white") +
  labs(title = "Per Capita GDP for 2007-2017",
     caption = "Source:https://data.worldbank.org")

# arrange the two plots above
library(patchwork)

plot1a + plot1b + plot_annotation(title = 
"Histograms for Per Capita Fixed Broadband Subscriptions and Per Capita GDP", 
                                tag_levels="a", tag_suffix = ")")

```

<br>

To further explore the distribution of the two variables, `gdpercap` and
`percapfbroad`, the filtered data based on individual year of observation will
be explored. Given there are 11 sets of filtered data, the `filter2007`,
`filtered2012`, and `filtered2017` datasets will be used as example to
demonstrate the distribution of the DV and IV. 

```{r}
# 2007 data
plot2a <- ggplot(data = filter2007, mapping = aes(x = percapfbroad)) +
  geom_histogram(fill = "aquamarine2", bins = 15, color = "white") +
  labs(title = "Per Capita Fixed Broadband 
       Subscriptions for 2007",
     caption = "Source:https://data.worldbank.org") 

plot2b <- ggplot(data = filter2007, mapping = aes(x = gdpercap)) +
  geom_histogram(fill = "bisque2", bins = 15, color = "white") +
  labs(title = "Per Capita GDP for 2007",
     caption = "Source:https://data.worldbank.org")

plot2a + plot2b + plot_annotation(title = "Histograms 2007 Data (`filter2007`)", 
                                tag_levels="a", tag_suffix = ")")


# 2012 data
plot3a <- ggplot(data = filter2012, mapping = aes(x = percapfbroad)) +
  geom_histogram(fill = "aquamarine4", bins = 15, color = "white") +
  labs(title = "Per Capita Fixed Broadband 
       Subscriptions for 2012",
     caption = "Source:https://data.worldbank.org") 

plot3b <- ggplot(data = filter2012, mapping = aes(x = gdpercap)) +
  geom_histogram(fill = "cornsilk4", bins = 15, color = "grey") +
  labs(title = "Per Capita GDP for 2012",
     caption = "Source:https://data.worldbank.org")

plot3a + plot3b + plot_annotation(title = "Histograms 2012 Data (`filter2012`)", 
                                tag_levels="a", tag_suffix = ")")


# 2017 data 
plot4a <- ggplot(data = filter2017, mapping = aes(x = percapfbroad)) +
  geom_histogram(fill = "aquamarine3", bins = 15, color = "white") +
  labs(title = "Per Capita Fixed Broadband 
       Subscriptions for 2017",
     caption = "Source:https://data.worldbank.org") 

plot4b <- ggplot(data = filter2017, mapping = aes(x = gdpercap)) +
  geom_histogram(fill = "cornsilk3", bins = 15, color = "grey") +
  labs(title = "Per Capita GDP for 2017",
     caption = "Source:https://data.worldbank.org")

plot4a + plot4b + plot_annotation(title = "Histograms 2017 Data (`filter2017`)", 
                                tag_levels="a", tag_suffix = ")")


```


***

## Visualization 

### Correlation and Regression Analysis 

In this anaysis, I'm examining the statistical relationship between `gdpercap`
(IV) and `percapfbroad` (DV). To further analyze how the two variables relate to
one another, the filtered dataset by year would be examined to explore how over
time the relationship between the two variable change. This will be useful at
not only answering the research question but also to demonstrate the effect of
time of the relationship between the IV and DV.

Furthermore, other variables including `percapelect` and `percapmobsub` will be
included in the some of the statistical models to examine whether these factors
have a moderator effect on the relationship between the IV and the DV.


This is the description of my null and alternative hypothesis:

* H~0~ (*null hypothesis*): High per capita GDP does not correspond to a high
per capita fixed broadband subscription in a country.

* H~1~ (*alternative hypothesis*): High per capita GDP does correspond to a high
per capita fixed broadband subscription in a country.


```{r}
options(scipen = 999, digits = 3)

# Correlation 
cor1 <- cor(myDataNew$gdpercap, myDataNew$percapfbroad, 
            use="pairwise.complete.obs")
cor1

# Bivariate Regression 
lm1 <- lm(percapfbroad ~ gdpercap, data = myDataNew)
summary(lm1)

# Plot the relationship
plot(percapfbroad ~ gdpercap,
       data = myDataNew, 
       main = "Scatterplot of per Capita Fixed Broadband Subscriptions and 
     per Capita GDP")

# add the regression line 
abline(lm1)

```


*For cor1*

* There is a positive correlation of **0.766**. That is there is a positive
relationship between per capita GDP and per capita fixed broadband subscription
across countries.

*For model lm(1)*

* The p-value is less than 0.05. That is, the relationship between country's per
capita GDP and per capita fixed broadband subscription is significant at 0.05
level. Therefore, there is evidence to reject the null hypothesis for the
alternative hypothesis at 0.05 level.

***
```{r}
# Test the correlation or the filtered dataset for 2007, 2009, 2011, 2013, 2015,
# and 2017

(cor2 <- cor(filter2007$gdpercap, filter2007$percapfbroad, 
            use="pairwise.complete.obs"))

(cor3 <- cor(filter2009$gdpercap, filter2009$percapfbroad, 
            use="pairwise.complete.obs"))

(cor4 <- cor(filter2011$gdpercap, filter2011$percapfbroad, 
            use="pairwise.complete.obs"))

(cor5 <- cor(filter2013$gdpercap, filter2013$percapfbroad, 
            use="pairwise.complete.obs"))

(cor6 <- cor(filter2015$gdpercap, filter2015$percapfbroad, 
            use="pairwise.complete.obs"))

(cor7 <- cor(filter2017$gdpercap, filter2017$percapfbroad, 
            use="pairwise.complete.obs"))

```

*For cor2*

* There is a positive correlation of **0.833**. That is there is a positive
relationship between per capita GDP and per capita fixed broadband subscription
across countries.


*For cor3*

* There is a positive correlation of **0.837**. That is there is a positive
relationship between per capita GDP and per capita fixed broadband subscription
across countries.


*For cor4*

* There is a positive correlation of **0.777**. That is there is a positive
relationship between per capita GDP and per capita fixed broadband subscription
across countries.


*For cor5*

* There is a positive correlation of **0.761**. That is there is a positive
relationship between per capita GDP and per capita fixed broadband subscription
across countries.


*For cor6*

* There is a positive correlation of **0.765**. That is there is a positive
relationship between per capita GDP and per capita fixed broadband subscription
across countries.


*For cor7*

* There is a positive correlation of **0.752**. That is there is a positive
relationship between per capita GDP and per capita fixed broadband subscription
across countries.

***

```{r}
# Regression for the filtered dataset for 2007, 2009, 2011, 2013, 2015, and 2017

lm2 <- lm(percapfbroad ~ gdpercap, data = filter2007)
summary(lm2)

lm3 <- lm(percapfbroad ~ gdpercap, data = filter2009)
summary(lm3)

lm4 <- lm(percapfbroad ~ gdpercap, data = filter2011)
summary(lm4)

lm5 <- lm(percapfbroad ~ gdpercap, data = filter2013)
summary(lm5)

lm6 <- lm(percapfbroad ~ gdpercap, data = filter2015)
summary(lm6)

lm7 <- lm(percapfbroad ~ gdpercap, data = filter2017)
summary(lm7)

```

```{r}
# Test whether there is correlation between DV, IV, and other IVS
lm8 <- lm(percapfbroad ~ gdpercap + percapelect, data = myDataNew)
summary(lm8)

lm9 <- lm(percapfbroad ~ gdpercap + percapelect + percapmobsub, data = myDataNew)
summary(lm9)

```


*For model lm(2) to model lm(7)*

* The p-value is less than 0.05. That is, the relationship between country's per
capita GDP and per capita fixed broadband subscription is significant at 0.05
level in the year 2007, 2009, 2011, 2013, 2015, 2017. Therefore, there is
evidence to reject the null hypothesis for the alternative hypothesis at 0.05
level.


*For model lm(8)* 

* `gdpercap`: The p-value is significant at 0.05 level.
* `percapelect`: The p-value of ( 0.0067) is not statistically significant at
0.05 level


*For model lm(9)*

* `gdpercap`: The p-value is significant at 0.05 level. 
* `percapmobsub`: The p-value is significant at 0.05 level. 
* `percapelect`: The p-value of (0.49090) is not statistically significant at
0.05 level


***


### GG-Plot Data Visualization 

```{r message=FALSE, warning=FALSE}
# Scatterplot for `percapfbroad` and `gdpercap`
plot5 <- filter2007 %>%
  ggplot(mapping = aes(x = gdpercap, y = percapfbroad)) +
  geom_point(color = "darkgoldenrod1") +
  geom_smooth(method = "lm", color = "black", size = 0.5) +
  labs(title = "per Capita GDP and per Capita Fixed 
       Broadband for 2007")


plot6 <- filter2011 %>%
  ggplot(mapping = aes(x = gdpercap, y = percapfbroad)) +
  geom_point(color = "darkolivegreen3") +
  geom_smooth(method = "lm", color = "black", size = 0.5) +
  labs(title = "per Capita GDP and per Capita Fixed 
       Broadband for 2011")


plot7 <- filter2015 %>%
  ggplot(mapping = aes(x = gdpercap, y = percapfbroad)) +
  geom_point(color = "darkgoldenrod") +
  geom_smooth(method = "lm", color = "black", size = 0.5) +
  labs(title = "per Capita GDP and per Capita Fixed 
       Broadband for 2015")

plot8 <- filter2017 %>%
  ggplot(mapping = aes(x = gdpercap, y = percapfbroad)) +
  geom_point(color = "darkseagreen4") +
  geom_smooth(method = "lm", color = "black", size = 0.5) +
  labs(title = "per Capita GDP and per Capita Fixed 
       Broadband for 2017")

# Arrange the plots 
plot5 + plot6 + plot_annotation(title = "Scatterplots for 2007 and 2011 Data", 
                                tag_levels="a", tag_suffix = ")",
                                caption = "Source:https://data.worldbank.org")

plot7 + plot8 + plot_annotation(title = "Scatterplots for 2015 and 2017 Data", 
                                tag_levels="a", tag_suffix = ")",
                                caption = "Source:https://data.worldbank.org")

```



***

## Reflection 

Going through the process of data collection and compilation not only helped me
to learn and improve my analytical skills but also taught me to appreciate the
work that goes behind any statistical analysis. Drawing from the dataset and
some of the variables, there were several constraints. First, I had to
reconstruct the dataset by compiling several datasets from the World Bank to
have all my variables of interest in a single dataset. Second, the variable
fixed broadband subscriptions included both residential and organizational
subscriptions. This poses some challenges in analyzing the distribution of those
subscriptions to understand how many households in a country had access to
high-speed internet at their places of residence.

Moving on to data wrangling and exploration, I encountered several challenges in
variable transformation and in making analytical choices around the handling of
observations across all 11 years. Given that I was using a panel dataset with
data from 2007 to 2017, I wish I was able to perform some more advanced
regression analysis to examine the effect of time on the relationship between my
IV and DV.

If I were to continue the project, I would examine the effect of fixed broadband
internet on countries' economic growth and development. The dataset used in this
project could also be useful in examining such an effect.


***

## Conclusion 

In brief, this paper examined whether countries' per capita GDP influences
people's access to fixed broadband internet. Data from 2007 to 2017 was used in
the examination of this research question. To analyze whether per capita GDP
(IV) had an influence on per capita fixed broadband subscriptions (DV) in a
country, I run a correlation between the two variables based on the entire
dataset with data from 2007 to 2017. I found a positive correlation of 0.776
between the IV and DV. I went further to examine the relationship between the
variable based on data from an individual year. Six correlations were run, and
they all revealed a positive coefficient, indicating a positive relationship
between per capita GDP and per capita fixed broadband subscription across the
six years included. There was an incremental change in the correlation
coefficient from 2007 to 2011, however, from 2011 to 2017 there was a
decremental change in the correlation coefficient. Given that correlation does
not mean causality, I went on to conduct linear regression analysis.

For regression analysis, I run 9 linear regression models. While models (lm1),
lm(8), and lm(9) used the large dataset with all 11 years of observations,
models lm(2) to lm(7) were based on the filtered dataset with 177 observations
in a particular year (2007, 2009, 2011, 2013, 2015, and 2017). For models lm(1)
to lm(7), the p-values were less than 0.05. That is, the relationship between a
country’s per capita GDP and per capita fixed broadband subscription was
statistically significant at 0.05 level in the years 2007, 2009, 2011, 2013,
2015, 2017 as well as the overall period between 2007 to 2017. Thus, there was
evidence to reject the null hypothesis for the alternative hypothesis.

In model lm(8), I regressed per capita fixed broadband subscription on per
capita GDP and per capita electricity. While the p-values for per capita GDP was
statistical significant at 0.05 level, that of per capita electricity was not
statistically significant. In model lm(9), I regressed per capita fixed
broadband subscription on per capita GDP, per capita electricity, and per capita
cellular mobile subscription. When the p-values for both per capita GDP and per
capita cellular mobile subscription were statistically significant at 0.05
level, the p-value for per capita electricity was not statistically significant
at 0.05 level. 

The results from all regression models are summarized in the table below.


### Model Summary 

```{r}
library(sjmisc)
library(sjPlot)
library(sjlabelled)

tab_model(lm1, lm2, lm3, lm4, lm5, lm6, lm7, lm8, lm9,
          title= "Table 1. Linear Panel Regression Models of Fixed Broadband 
          Internet Access",
          dv.labels = c("lm1", "lm2", "lm3", "lm4", "lm5", "lm6", 
                        "lm7", "lm8", "lm9"),
          string.p = "P-Value",
          show.ci = FALSE)

 
```


***

## Reference


Alam, T. F., Sultana, N., & Rayhan, M. I. (2019). Structural equation modeling:
An application of broadband penetration and GDP growth in Asia. Journal of
Economic Structures, 8(1), 1–11.
https://ideas.repec.org/a/spr/jecstr/v8y2019i1d10.1186_s40008-019-0148-0.html

Barry, J. (n.d.). COVID-19 exposes why access to the internet is a human right.
OpenGlobalRights.
https://www.openglobalrights.org/covid-19-exposes-why-access-to-internet-is-human-right/

Hadley Wickham, Romain François, Lionel Henry and Kirill Müller (2021). dplyr: A
Grammar of Data Manipulation. R package version 1.0.7.
https://CRAN.R-project.org/package=dplyr

International Telecommunication Union. (2017). ICT Facts and Figures 2017.
https://about.fb.com/wp-content/uploads/2018/08/ICTFactsFigures2017.pdf

Khandelwal, Saloni & Agarwal, Sonali. (n.d.). Application Of Fixed Broadband
Subscriptions Impact On Economic Growth.
https://doi.org/10.21203/rs.3.rs-60231/v1

Lüdecke D (2018). “sjmisc: Data and Variable Transformation Functions.” _Journal
of Open Source Software_, *3*(26), 754. doi: 10.21105/joss.00754 (URL:
https://doi.org/10.21105/joss.00754).
  
Lüdecke D (2021). _sjPlot: Data Visualization for Statistics in Social Science_.
R package version 2.8.9, <URL: https://CRAN.R-project.org/package=sjPlot>.

Lüdecke D (2021). _sjlabelled: Labelled Data Utility Functions (Version 1.1.8)_.
doi: 10.5281/zenodo.1249215 (URL: https://doi.org/10.5281/zenodo.1249215), <URL:
https://CRAN.R-project.org/package=sjlabelled>.

Lüdecke, D. (2021, July 10). Summary of Regression Models as HTML Table.
https://cran.r-project.org/web/packages/sjPlot/vignettes/tab_model_estimates.html

McDonough, Carol C. (2012, July 1). Fixed and mobile broadband: Demand and
market structure. 23rd European Regional Conference of the International
Telecommunications Society (ITS), Vienna, Austria.
https://www.econstor.eu/bitstream/10419/60350/1/720272319.pdf

Minges, M. (2015). Digital Dividends: Exploring the Relationship Between
Broadband and Economic Growth.
https://openknowledge.worldbank.org/bitstream/handle/10986/23638/WDR16-BP-Exploring-the-Relationship-between-Broadband-and-Economic-Growth-Minges.pdf;sequence=1

Poushter, J. (2016, February 22). Internet access grows worldwide but remains
higher in advanced economies. Pew Research Center.
https://www.pewresearch.org/global/2016/02/22/internet-access-growing-worldwide-but-remains-higher-in-advanced-economies/

Revelle, W. (2021) psych: Procedures for Personality and Psychological
Research, Northwestern University, Evanston, Illinois, USA,
https://CRAN.R-project.org/package=psych Version = 2.1.6,.

Wempen, F. (2015). Different Types of Internet Connections—Dummies.
https://www.dummies.com/education/internet-basics/different-types-of-internet-connections/

Wickham, Hadley & Grolemund, Garrett. (2017). R for Data Science.
https://r4ds.had.co.nz/index.html

Wickham et al., (2019). Welcome to the tidyverse. Journal of Open Source
Software, 4(43), 1686, https://doi.org/10.21105/joss.01686

World Bank. (n.d). Fixed broadband subscriptions | Data Catalog. https://datacatalog.worldbank.org/fixed-broadband-subscriptions-1 








